---
title: "Assignment 2: Coding Basics"
author: Lily Zhang
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. I use sequence to generate numbers. The first two numbers in parentheses mean the range of the sequence, the third number means increasing by three. I assign the sequence as A. 
A <- seq(1,30,3)
#2. I compute the mean and median of sequence A and assign mean as mean_A, median as median_A.
mean_A <- mean(A)
median_A <- median(A)
#3. I test if mean_A is greater than median_A and the result is FALSE, which indicates mean is not greater than the median. 
mean_A > median_A
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#I assigned a collection of five student names of my friends.
student_name <- c("Lily", "Hope", "Mia", "Christal", "Zhuowen")
#I used this function to randomly generated five integer within the range of 0 to 100, corresponding to names above.
test_score <- sample.int(100,5)
#I test if the number generated above is larger or equal to 50. If so, it returns as 'TRUE', if not, it returns as 'FALSE.'
pass_notpass <- test_score >=50
#I constructed a data frame titled as test_result, encompassing the previous information.
test_result <- data.frame (student_name, test_score, pass_notpass)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frame can consist of different data types in each column but matrix can only contain numerical data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#I apply an 'ifelse' function that tests whether if the given value equals or exceeds 50 (the standard of passing the test). If yes, the result prints as pass, and if not, the result prints as fail. The ifelse function is named as test.
test <- function (x) {
  result <- ifelse(x >= 50,"pass","fail")
  print(result)
}
#I assigned the function test with value of test_score generated in section 2. 
test(test_score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' works for the previous question because it works well with vectors and data frames while 'if' and 'else' are used strictly for scalar.
